\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{geometry}
\geometry{margin=1in}

\title{Database Schema and Query Descriptions \\ UIBK - Music Streaming Service}
\author{Prof. Fedeli Massimo - IIS Fermi Sacconi Cpia}
\date{}

\begin{document}
	
	\maketitle
	
	\section*{Schema Definition}
	
	The database contains the following relations:
	
	\begin{itemize}
		\item \textbf{Artist}(\textit{ArtistId}, \textit{Name})
		\item \textbf{Album}(\textit{AlbumId}, \textit{Title}, \textit{ArtistId})
		\item \textbf{Track}(\textit{TrackId}, \textit{Name}, \textit{AlbumId}, \textit{GenreId}, \textit{Milliseconds}, \textit{Bytes}, \textit{UnitPrice})
		\item \textbf{PlaylistContent}(\textit{PlaylistId}, \textit{TrackId})
		\item \textbf{Playlist}(\textit{PlaylistId}, \textit{Name})
		\item \textbf{Genre}(\textit{GenreId}, \textit{Name})
		\item \textbf{Invoice}(\textit{InvoiceId}, \textit{CustomerId}, \textit{InvoiceDate}, \textit{Total})
		\item \textbf{InvoiceParts}(\textit{InvoicePartId}, \textit{InvoiceId}, \textit{TrackId}, \textit{UnitPrice}, \textit{Quantity})
		\item \textbf{Customer}(\textit{CustomerId}, \textit{FirstName}, \textit{LastName}, \textit{Address}, \textit{City}, \textit{Country}, \textit{PostalCode}, \textit{Email})
	\end{itemize}
	
	\section*{Queries (Natural Language)}
	
	\begin{enumerate}
		\item Find the name of all artists.
		\item Find the title of all albums by the artist with ArtistId = 1.
		\item Find the name of all tracks belonging to the "Rock" genre.
		\item Find the name of all tracks included in the playlist with PlaylistId = 5.
		\item Find the name of all customers who have purchased at least one track.
		\item Find the full name (FirstName + LastName) of all customers residing in Italy.
		\item Find the total number of tracks in the database.
		\item Find the name of all playlists that contain tracks of the "Jazz" genre.
		\item Find the name of all tracks that have been purchased at least once.
		\item Find the average unit price of all tracks.
		\item Find the name of all artists who have released at least one album.
		\item Find the name of all tracks with a duration longer than 300,000 milliseconds.
		\item Find the name of all customers who made a purchase on 2026-01-11.
		\item Find the name of all tracks included in any playlist.
		\item Find the name of all tracks that have never been purchased.
		\item Find the name of all artists who have not released any album.
		\item Find the name of all playlists containing more than 10 tracks.
		\item Find the name of all tracks in the "Classical" genre with a unit price lower than 1.00.
		\item Find the name of all customers who have spent more than 10 euros in total.
		\item Find the name of all tracks that belong to albums by Italian artists (assuming an artist is Italian if their name contains "Italy"—for example only).
		\item Find the name of all tracks purchased by customers residing in Vienna.
		\item Find the name of all playlists that contain tracks from more than one genre.
		\item Find the name of all tracks that appear in at least two different playlists.
		\item Find the name of all customers who have purchased tracks from every genre present in the database.
		\item Find the name of all tracks that have never been included in any playlist.
	\end{enumerate}
	
\end{document}